---
title: "subsppLabelR_2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setting Up the Package

First we must install the R package before we can use it. 

```{r install}
library(devtools)
devtools::install_github('kaiyaprovost/subsppLabelR',force=F)
library(subsppLabelR)
```

## Niche Calculations

Now lets start using these occurence data. First we will download some WorldClim data. For the sake of this example, we will only use bio1 and bio12, which are mean annual temperature and mean annual precipitation. We will also crop the data to match our previous data -- but remember to change your cropping based on your own ENMs you wish to calculate. 

```{r worldclim}
wcdata = raster::getData("worldclim",var="bio",res=10)
wcdata = wcdata[[c(1,12)]]
wcdata = crop(wcdata,extent(bgLayer))
plot(wcdata)

```

Next we will remove any points that do not have associated environmental data and thin the data. We will only thin the data once, but ideally you should thin multiple times. This will take some time. 

```{r thinpoints}

cleanByEnvironment = function(Env,loc){
  loc[,2] = as.numeric(loc[,2])
  loc[,3] = as.numeric(loc[,3])
  extr = raster::extract(Env, loc[,2:3]) ## gets values from Env that are at loc
  head(extr)
  loc_clean = loc[!is.na(extr[,1]),]
  print(paste("Removed",nrow(loc)-nrow(loc_clean),"rows with no Env data"))
  
  return(loc_clean)
}

loc_good_clean = cleanByEnvironment(Env=wcdata,loc=loc_good)

spThinBySubspecies = function(loc_good_clean,thin.par=10,reps=1,lat.col="latitude",
                              long.col="longitude",spec.col="assigned"){
  
  locs_thinned=lapply(unique(loc_good_clean$assigned),FUN=function(subspp){
    print(subspp)
    loc_temp = loc_good_clean[loc_good_clean$assigned==subspp,]
    
    ## note: if you do not change the "name" column, it will error out and only use the first subspecies. 
    loc_thin = spThin::thin(loc.data = loc_temp,
                            lat.col = lat.col,
                            long.col = long.col,
                            spec.col = spec.col,
                            thin.par = thin.par, ## km distance that records need to be separated by
                            reps = reps, ## number of times to repeat thinning process
                            locs.thinned.list.return = T,
                            write.files = F,
                            max.files = 1,
                            write.log.file = F)[[1]]
    loc_thin$assigned = subspp
    return(loc_thin)
  }) 
  
  
  loc_thin = do.call(rbind,locs_thinned)
  return(loc_thin)
}

loc_thin = spThinBySubspecies(loc_good_clean,thin.par=10,reps=1,lat.col="latitude",
                              long.col="longitude",spec.col="assigned")

```

Now we have our datasets. Let us calculate our background variation. 

```{r backgroundpca}

backgroundForPCA = function(localities=loc_good[,c("Longitude","Latitude")],
                            r=200000,
                            num=(100*nrow(localities)),
                            e=Env){
  library(ENMTools)
  
  bg1 = ENMTools::background.points.buffer(localities, radius = r,
                                           n = num, mask = e[[1]])
  extract1 = na.omit(cbind(localities,
                           extract(e, localities), rep(1, nrow(localities))))
  colnames(extract1)[ncol(extract1)] = 'occ'
  extbg1 = na.omit(cbind(bg1, extract(e, bg1), rep(0, nrow(bg1))))
  colnames(extbg1)[ncol(extbg1)] = 'occ'
  dat1 = rbind(extract1, extbg1)
  return(list(bgenv=dat1,bgpoints=bg1,bgext=extract1,bgextbg=extbg1))
}

loc_thin_bgstuff = backgroundForPCA(localities=loc_thin[,c("Longitude","Latitude")],
                                    num=20000,
                                    e=wcdata)
bg_dat = loc_thin_bgstuff$bgenv
bg_bg = loc_thin_bgstuff$bgpoints

## and per taxon
backgroundPerSpecies = function(localities=loc_thin,e,num){
  loc_thin_by_subspecies = split(loc_thin, loc_thin$assigned)
  bgenv_by_subspecies = list()
  
  bgext_by_subspecies = list()
  bgpoints_by_subspecies = list()
  bgextbg_by_subspecies = list()
  
  for(i in 1:length(names(loc_thin_by_subspecies))){
    singleSubspp = loc_thin_by_subspecies[[i]]
    subsppName = names(loc_thin_by_subspecies)[[i]]
    single_bgstuff = backgroundForPCA(singleSubspp[,c("Longitude","Latitude")],e=e,num=num)
    bgenv_by_subspecies[[i]] = single_bgstuff$bgenv
    bgext_by_subspecies[[i]] = single_bgstuff$bgext
    bgpoints_by_subspecies[[i]] = single_bgstuff$bgpoints
    bgextbg_by_subspecies[[i]] = single_bgstuff$bgextbg
    
  }
  names(bgenv_by_subspecies) = names(loc_thin_by_subspecies)
  names(bgext_by_subspecies) = names(loc_thin_by_subspecies)
  names(bgpoints_by_subspecies) = names(loc_thin_by_subspecies)
  names(bgextbg_by_subspecies) = names(loc_thin_by_subspecies)
  
  return(list(bgenv_by_subspecies=bgenv_by_subspecies,
              bgext_by_subspecies=bgext_by_subspecies,
              bgpoints_by_subspecies=bgpoints_by_subspecies,
              bgextbg_by_subspecies=bgextbg_by_subspecies))
}

perspecies_bgstuff = backgroundPerSpecies(localities=loc_thin,e=wcdata,num=20000)




```

Time to run some functions on the niche. First, we run a custom ecospat function that handles NA values as well as a function that compares the PCA values in each subspecies' home range. 

```{r ecospatstuff}

ecospat.grid.clim.dyn_custom <- function(glob, glob1, sp, R, th.sp = 0, th.env = 0,
                                         geomask = NULL,removeNA=T) {
  
  glob <- as.matrix(glob)
  glob1 <- as.matrix(glob1)
  sp <- as.matrix(sp)
  l <- list()
  
  if (ncol(glob) > 2)
    stop("cannot calculate overlap with more than two axes")
  
  if (ncol(glob) == 1) {
    # if scores in one dimension (e.g. LDA,SDM predictions,...)
    xmax <- max(glob[, 1])
    xmin <- min(glob[, 1])
    x <- seq(from = min(glob[, 1]), to = max(glob[, 1]), length.out = R) # breaks on score gradient 1
    sp.dens <- density(sp[, 1], kernel = "gaussian", from = xmin, to = xmax,
                       n = R, cut = 0) # calculate the density of occurrences in a vector of R pixels along the score gradient
    # using a gaussian kernel density function, with R bins.
    glob1.dens <- density(glob1[, 1], kernel = "gaussian", from = xmin,
                          to = xmax, n = R, cut = 0) # calculate the density of environments in glob1
    z <- sp.dens$y * nrow(sp)/sum(sp.dens$y) # rescale density to the number of occurrences in sp
    # number of occurrence/pixel
    Z <- glob1.dens$y * nrow(glob)/sum(glob1.dens$y) # rescale density to the number of sites in glob1
    glob1r <- sapply(glob1, findInterval, glob1.dens$x)
    th.env <- quantile(glob1.dens$y[glob1r], th.env)
    glob1rm <- which(Z < th.env)
    spr <- sapply(sp, findInterval, sp.dens$x)
    th.sp <- quantile(sp.dens$y[spr], th.sp)
    sprm <- which(z < th.sp)
    z[sprm] <- 0 # remove infinitesimally small number generated by kernel density function
    Z[glob1rm] <- 0 # remove infinitesimally small number generated by kernel density function
    
    z.uncor <- z/max(z) # rescale between [0:1] for comparison with other species
    z.cor <- z/Z # correct for environment prevalence
    z.cor[is.na(z.cor)] <- 0 # remove n/0 situations
    z.cor[z.cor == "Inf"] <- 0 # remove 0/0 situations
    z.cor <- z.cor/max(z.cor) # rescale between [0:1] for comparison with other species
    w <- z.uncor
    w[w > 0] <- 1
    l$x <- x
    l$z <- z
    l$z.uncor <- z.uncor
    l$z.cor <- z.cor
    l$Z <- Z
    l$glob <- glob
    l$glob1 <- glob1
    l$sp <- sp
    l$w <- w
  }
  
  if (ncol(glob) == 2) {
    # if scores in two dimensions (e.g. PCA)
    
    xmin <- min(glob[, 1])
    xmax <- max(glob[, 1])
    ymin <- min(glob[, 2])
    ymax <- max(glob[, 2]) # data preparation
    glob1r <- data.frame(cbind((glob1[, 1] - xmin)/abs(xmax - xmin), (glob1[,
                                                                            2] - ymin)/abs(ymax - ymin))) # data preparation
    spr <- data.frame(cbind((sp[, 1] - xmin)/abs(xmax - xmin), (sp[, 2] -
                                                                  ymin)/abs(ymax - ymin))) # data preparation
    mask <- adehabitatMA::ascgen(SpatialPoints(cbind((0:(R))/R, (0:(R)/R))),
                                 nrcol = R-2, count = FALSE) # data preparation
    sp.dens <- adehabitatHR::kernelUD(SpatialPoints(spr[, 1:2]), h = "href", grid = mask,
                                      kern = "bivnorm") # calculate the density of occurrences in a grid of RxR pixels along the score gradients
    sp.dens <- raster(xmn = xmin, xmx = xmax, ymn = ymin, ymx = ymax, matrix(sp.dens$ud,
                                                                             nrow = R))
    # using a gaussian kernel density function, with RxR bins.
    # sp.dens$var[sp.dens$var>0 & sp.dens$var<1]<-0
    glob1.dens <- adehabitatHR::kernelUD(SpatialPoints(glob1r[, 1:2]), grid = mask, kern = "bivnorm")
    glob1.dens <- raster(xmn = xmin, xmx = xmax, ymn = ymin, ymx = ymax,
                         matrix(glob1.dens$ud, nrow = R))
    # glob1.dens$var[glob1.dens$var<1 & glob1.dens$var>0]<-0
    
    x <- seq(from = min(glob[, 1]), to = max(glob[, 1]), length.out = R) # breaks on score gradient 1
    y <- seq(from = min(glob[, 2]), to = max(glob[, 2]), length.out = R) # breaks on score gradient 2
    glob1r <- extract(glob1.dens, glob1)
    Z.th <- quantile(glob1r, th.env,na.rm=removeNA)
    glob1.dens[glob1.dens < Z.th] <- 0
    if (!is.null(geomask)) {
      proj4string(geomask) <- NA
      glob1.dens <- mask(glob1.dens, geomask, updatevalue = 0) # Geographical mask in the case if the analysis takes place in the geographical space
    }
    Z <- glob1.dens * nrow(glob1)/cellStats(glob1.dens, "sum")
    
    spr <- extract(sp.dens, sp)
    z.th <- quantile(spr, th.sp)
    sp.dens[Z == 0] <- 0
    sp.dens[sp.dens < z.th] <- 0
    if (!is.null(geomask)) {
      sp.dens <- mask(sp.dens, geomask, updatevalue = 0) # Geographical mask in the case if the analysis takes place in the geographical space
    }
    z <- sp.dens * nrow(sp)/cellStats(sp.dens, "sum")
    z.uncor <- z/cellStats(z, "max")
    w <- z.uncor # remove infinitesimally small number generated by kernel density function
    w[w > 0] <- 1
    z.cor <- z/Z # correct for environment prevalence
    z.cor[is.na(z.cor)] <- 0 # remove n/0 situations
    z.cor <- z.cor/cellStats(z.cor, "max")
    l$x <- x
    l$y <- y
    l$z <- z
    l$z.uncor <- z.uncor
    l$z.cor <- z.cor
    l$Z <- Z
    l$glob <- glob
    l$glob1 <- glob1
    l$sp <- sp
    l$w <- w
    
  }
  
  return(l)
}

createPcaToCompare = function(loc_thin_bgstuff,perspecies_bgstuff,species) {
  bg_dat = loc_thin_bgstuff$bgenv
  bg_bg = loc_thin_bgstuff$bgpoints
  
  bgext_by_subspecies = perspecies_bgstuff$bgext_by_subspecies
  bgenv_by_subspecies = perspecies_bgstuff$bgenv_by_subspecies
  
  ## pca bg points
  pca.env <- ade4::dudi.pca(bg_dat[,3:(ncol(bg_dat)-1)],scannf=F,nf=2)
  #png(paste("PCAcorrelationCircle_",species,".png",sep=""))
  ecospat::ecospat.plot.contrib(contrib=pca.env$co, eigen=pca.env$eig)
  #dev.off()
  
  ## pca scores whole study area, all points, all subspecies
  scores_globclim<-pca.env$li # PCA scores for the whole study area (all points)
  
  ## now get pca scores per species instead
  
  scores = list()
  scores_clim = list()
  grid_clim = list()
  
  for(i in 1:length(names(bgext_by_subspecies))){
    singleSubspp_bgext = bgext_by_subspecies[[i]]
    singleSubspp_bgenv = bgenv_by_subspecies[[i]]
    subsppName = names(bgext_by_subspecies)[[i]]
    
    scores_subspp = ade4::suprow(pca.env,
                                 singleSubspp_bgext[which(
                                   singleSubspp_bgext[,ncol(singleSubspp_bgext)]==1)
                                   ,3:(ncol(singleSubspp_bgext)-1)])$li # PCA scores for the species 1 distribution
    
    
    scores[[i]] = scores_subspp
    
    scores_clim_subspp = ade4::suprow(pca.env,
                                      singleSubspp_bgenv[,3:(ncol(singleSubspp_bgenv)-1)])$li # PCA scores for the whole native study area species 1 ## bgenv
    
    scores_clim[[i]] = scores_clim_subspp
    
    ## make a dynamic occurrence densities grid
    ## grid of occ densities along one or two environmental gradients
    ## glob = env variables in background
    ## glob1 = env variables for species
    ## sp = occurrences of species
    ## R = resolution
    ## th.sp = a threshhold to elimite low density values of species occurrences
    grid_clim_subspp <- ecospat.grid.clim.dyn_custom(glob = scores_globclim,
                                                     glob1 = scores_clim_subspp,
                                                     sp = scores_subspp,
                                                     R = 100,
                                                     th.sp = 0,
                                                     th.env = 0,
                                                     removeNA=T)
    grid_clim[[i]] = grid_clim_subspp
    
  }
  
  names(scores) = names(bgext_by_subspecies)
  names(scores_clim) = names(bgext_by_subspecies)
  names(grid_clim) = names(bgext_by_subspecies)
  
  #pdf(paste("NicheSpaceComparison_",species,".pdf",sep=""))
  par(mfrow=n2mfrow(length(grid_clim)),
      ask=F)
  for(i in 1:length(grid_clim)){
    plot(grid_clim[[i]]$w,main=names(grid_clim)[[i]])
  }
  #dev.off()
  
  return(list(scores_globclim=scores_globclim,
              scores=scores,
              scores_clim=scores_clim,
              grid_clim=grid_clim))
  
}


pcaOutput = createPcaToCompare(loc_thin_bgstuff=loc_thin_bgstuff,
                               perspecies_bgstuff=perspecies_bgstuff,
                               species=species)


pca_grid_clim = pcaOutput$grid_clim

```


Now we calculate the pairwise niche overlap of the taxa. 

```{r pairwiseniche}

pairwiseNicheOverlap = function(pca_grid_clim=pca_grid_clim){
  
  overlap_df = data.frame(spp1=character(),
                          spp2=character(),
                          SchoenersD=numeric(),
                          modifiedHellingersI=numeric())
  
  for(i in 1:length(pca_grid_clim)){
    for(j in 1:length(pca_grid_clim)){
      if(i<j){
        #print(paste(i,j))
        spp1_name = names(pca_grid_clim)[[i]]
        spp1 = pca_grid_clim[[i]]
        spp2_name = names(pca_grid_clim)[[j]]
        spp2 = pca_grid_clim[[j]]
        overlap <- ecospat::ecospat.niche.overlap(spp1, spp2, cor=T)
        rowToAdd = cbind(as.character(spp1_name),
                         as.character(spp2_name),
                         as.numeric(overlap$D),
                         as.numeric(overlap$I))
        colnames(rowToAdd) = colnames(overlap_df)
        overlap_df = rbind(overlap_df,rowToAdd)
      }
    }
  }
  return(overlap_df)
}
overlap_df = pairwiseNicheOverlap(pca_grid_clim=pca_grid_clim)
print(overlap_df)

```

And we test for niche equivalency in the taxa. Note that this requires a few modifications to ecospat code. This involves simulations to approximate a p-value: this process can take quite a while; feel free to reduce the amount of replications it goes through by changing "rep1" or "rep2". 

```{r nicheeq}
## test for niche equvalence pairwise
## first need to modify function again to remove NA
ecospat.niche.equivalency.test_custom <- function(z1, z2, rep, alternative = "greater", ncores=1) {
  
  R <- length(z1$x)
  l <- list()
  
  obs.o <- ecospat::ecospat.niche.overlap(z1, z2, cor = TRUE) #observed niche overlap
  
  if (ncores == 1){
    sim.o <- as.data.frame(matrix(unlist(lapply(1:rep, overlap.eq.gen_custom, z1, z2)), byrow = TRUE,
                                  ncol = 2)) #simulate random overlap
  }else{
    #number of cores attributed for the permutation test
    cl <- makeCluster(ncores) #open a cluster for parallelization
    invisible(clusterEvalQ(cl)) #import the internal function into the cluster
    sim.o <- as.data.frame(matrix(unlist(parLapply(cl, 1:rep, overlap.eq.gen_custom, z1, z2)), byrow = TRUE,
                                  ncol = 2)) #simulate random overlap
    stopCluster(cl) #shutdown the cluster
  }
  colnames(sim.o) <- c("D", "I")
  l$sim <- sim.o # storage
  l$obs <- obs.o # storage
  
  if (alternative == "greater") {
    l$p.D <- (sum(sim.o$D >= obs.o$D) + 1)/(length(sim.o$D) + 1) # storage of p-values alternative hypothesis = greater -> test for niche conservatism/convergence
    l$p.I <- (sum(sim.o$I >= obs.o$I) + 1)/(length(sim.o$I) + 1) # storage of p-values alternative hypothesis = greater -> test for niche conservatism/convergence
  }
  if (alternative == "lower") {
    l$p.D <- (sum(sim.o$D <= obs.o$D) + 1)/(length(sim.o$D) + 1) # storage of p-values alternative hypothesis = lower -> test for niche divergence
    l$p.I <- (sum(sim.o$I <= obs.o$I) + 1)/(length(sim.o$I) + 1) # storage of p-values alternative hypothesis = lower -> test for niche divergence
  }
  
  return(l)
}
#### internal functions from ENMtools that are needed for custom one
overlap.sim.gen <- function(repi, z1, z2, rand.type = rand.type) {
  R1 <- length(z1$x)
  R2 <- length(z2$x)
  if (is.null(z1$y) & is.null(z2$y)) {
    if (rand.type == 1) {
      # if rand.type = 1, both z1 and z2 are randomly shifted, if rand.type =2, only z2 is randomly
      # shifted
      center.z1 <- which(z1$z.uncor == 1) # define the centroid of the observed niche
      Z1 <- z1$Z/max(z1$Z)
      rand.center.z1 <- sample(1:R1, size = 1, replace = FALSE, prob = Z1) # randomly (weighted by environment prevalence) define the new centroid for the niche
      xshift.z1 <- rand.center.z1 - center.z1 # shift on x axis
      z1.sim <- z1
      z1.sim$z <- rep(0, R1) # set intial densities to 0
      for (i in 1:length(z1$x)) {
        i.trans.z1 <- i + xshift.z1
        if (i.trans.z1 > R1 | i.trans.z1 < 0)
          (next)() # densities falling out of the env space are not considered
        z1.sim$z[i.trans.z1] <- z1$z[i] # shift of pixels
      }
      z1.sim$z <- (z1$Z != 0) * 1 * z1.sim$z # remove densities out of existing environments
      z1.sim$z.cor <- (z1.sim$z/z1$Z)/max((z1.sim$z/z1$Z), na.rm = TRUE) #transform densities into occupancies
      z1.sim$z.cor[which(is.na(z1.sim$z.cor))] <- 0
      z1.sim$z.uncor <- z1.sim$z/max(z1.sim$z, na.rm = TRUE)
      z1.sim$z.uncor[which(is.na(z1.sim$z.uncor))] <- 0
    }
    
    center.z2 <- which(z2$z.uncor == 1) # define the centroid of the observed niche
    Z2 <- z2$Z/max(z2$Z)
    rand.center.z2 <- sample(1:R2, size = 1, replace = FALSE, prob = Z2) # randomly (weighted by environment prevalence) define the new centroid for the niche
    
    xshift.z2 <- rand.center.z2 - center.z2 # shift on x axis
    z2.sim <- z2
    z2.sim$z <- rep(0, R2) # set intial densities to 0
    for (i in 1:length(z2$x)) {
      i.trans.z2 <- i + xshift.z2
      if (i.trans.z2 > R2 | i.trans.z2 < 0)
        (next)() # densities falling out of the env space are not considered
      z2.sim$z[i.trans.z2] <- z2$z[i] # shift of pixels
    }
    z2.sim$z <- (z2$Z != 0) * 1 * z2.sim$z # remove densities out of existing environments
    z2.sim$z.cor <- (z2.sim$z/z2$Z)/max((z2.sim$z/z2$Z), na.rm = TRUE) #transform densities into occupancies
    z2.sim$z.cor[which(is.na(z2.sim$z.cor))] <- 0
    z2.sim$z.uncor <- z2.sim$z/max(z2.sim$z, na.rm = TRUE)
    z2.sim$z.uncor[which(is.na(z2.sim$z.uncor))] <- 0
  }
  
  if (!is.null(z2$y) & !is.null(z1$y)) {
    if (rand.type == 1) {
      # if rand.type = 1, both z1 and z2 are randomly shifted, if rand.type =2, only z2 is randomly
      # shifted
      centroid.z1 <- which(z1$z.uncor == 1, arr.ind = TRUE)[1, ] # define the centroid of the observed niche
      Z1 <- z1$Z/max(z1$Z)
      rand.centroids.z1 <- which(Z1 > 0, arr.ind = TRUE) # all pixels with existing environments in the study area
      weight.z1 <- Z1[Z1 > 0]
      rand.centroid.z1 <- rand.centroids.z1[sample(1:nrow(rand.centroids.z1), size = 1, replace = FALSE,
                                                   prob = weight.z1), ] # randomly (weighted by environment prevalence) define the new centroid for the niche
      xshift.z1 <- rand.centroid.z1[1] - centroid.z1[1] # shift on x axis
      yshift.z1 <- rand.centroid.z1[2] - centroid.z1[2] # shift on y axis
      z1.sim <- z1
      z1.sim$z <- matrix(rep(0, R1 * R1), ncol = R1, nrow = R1) # set intial densities to 0
      for (i in 1:R1) {
        for (j in 1:R1) {
          i.trans.z1 <- i + xshift.z1
          j.trans.z1 <- j + yshift.z1
          if (i.trans.z1 > R1 | i.trans.z1 < 0)
            (next)() # densities falling out of the env space are not considered
          if (j.trans.z1 > R1 | j.trans.z1 < 0)
            (next)()
          z1.sim$z[i.trans.z1, j.trans.z1] <- z1$z[i, j] # shift of pixels
        }
      }
      z1.sim$z <- (z1$Z != 0) * 1 * z1.sim$z # remove densities out of existing environments
      z1.sim$z.cor <- (z1.sim$z/z1$Z)/max((z1.sim$z/z1$Z), na.rm = TRUE) #transform densities into occupancies
      z1.sim$z.cor[which(is.na(z1.sim$z.cor))] <- 0
      z1.sim$z.uncor <- z1.sim$z/max(z1.sim$z, na.rm = TRUE)
      z1.sim$z.uncor[which(is.na(z1.sim$z.uncor))] <- 0
    }
    centroid.z2 <- which(z2$z.uncor == 1, arr.ind = TRUE)[1, ] # define the centroid of the observed niche
    Z2 <- z2$Z/max(z2$Z)
    rand.centroids.z2 <- which(Z2 > 0, arr.ind = TRUE) # all pixels with existing environments in the study area
    weight.z2 <- Z2[Z2 > 0]
    rand.centroid.z2 <- rand.centroids.z2[sample(1:nrow(rand.centroids.z2), size = 1, replace = FALSE,
                                                 prob = weight.z2), ] # randomly (weighted by environment prevalence) define the new centroid for the niche
    xshift.z2 <- rand.centroid.z2[1] - centroid.z2[1] # shift on x axis
    yshift.z2 <- rand.centroid.z2[2] - centroid.z2[2] # shift on y axis
    z2.sim <- z2
    z2.sim$z <- matrix(rep(0, R2 * R2), ncol = R2, nrow = R2) # set intial densities to 0
    for (i in 1:R2) {
      for (j in 1:R2) {
        i.trans.z2 <- i + xshift.z2
        j.trans.z2 <- j + yshift.z2
        if (i.trans.z2 > R2 | i.trans.z2 < 0)
          (next)() # densities falling out of the env space are not considered
        if (j.trans.z2 > R2 | j.trans.z2 < 0)
          (next)()
        z2.sim$z[i.trans.z2, j.trans.z2] <- z2$z[i, j] # shift of pixels
      }
    }
    z2.sim$z <- (z2$Z != 0) * 1 * z2.sim$z # remove densities out of existing environments
    z2.sim$z.cor <- (z2.sim$z/z2$Z)/max((z2.sim$z/z2$Z), na.rm = TRUE) #transform densities into occupancies
    z2.sim$z.cor[which(is.na(z2.sim$z.cor))] <- 0
    z2.sim$z.uncor <- z2.sim$z/max(z2.sim$z, na.rm = TRUE)
    z2.sim$z.uncor[which(is.na(z2.sim$z.uncor))] <- 0
  }
  
  if (rand.type == 1) {
    o.i <- ecospat::ecospat.niche.overlap(z1.sim, z2.sim, cor = TRUE)
  }
  if (rand.type == 2)
  {
    o.i <- ecospat::ecospat.niche.overlap(z1, z2.sim, cor = TRUE)
  } # overlap between random and observed niches
  sim.o.D <- o.i$D # storage of overlaps
  sim.o.I <- o.i$I
  return(c(sim.o.D, sim.o.I))
}
overlap.eq.gen_custom <- function(repi, z1, z2) {
  if (is.null(z1$y)) {
    # overlap on one axis
    
    occ.pool <- c(z1$sp, z2$sp) # pool of random occurrences
    rand.row <- sample(1:length(occ.pool), length(z1$sp)) # random reallocation of occurrences to datasets
    sp1.sim <- occ.pool[rand.row]
    sp2.sim <- occ.pool[-rand.row]
  }
  
  if (!is.null(z1$y)) {
    # overlap on two axes
    
    occ.pool <- rbind(z1$sp, z2$sp) # pool of random occurrences
    row.names(occ.pool)<-c() # remove the row names
    rand.row <- sample(1:nrow(occ.pool), nrow(z1$sp)) # random reallocation of occurrences to datasets
    sp1.sim <- occ.pool[rand.row, ]
    sp2.sim <- occ.pool[-rand.row, ]
  }
  
  z1.sim <- ecospat.grid.clim.dyn_custom(z1$glob, z1$glob1, data.frame(sp1.sim), R = length(z1$x)) # gridding
  z2.sim <- ecospat.grid.clim.dyn_custom(z2$glob, z2$glob1, data.frame(sp2.sim), R = length(z2$x))
  
  o.i <- ecospat::ecospat.niche.overlap(z1.sim, z2.sim, cor = TRUE) # overlap between random and observed niches
  sim.o.D <- o.i$D # storage of overlaps
  sim.o.I <- o.i$I
  return(c(sim.o.D, sim.o.I))
}

pairwiseNicheEquivalence = function(pca_grid_clim=pca_grid_clim,rep1=10,rep2=10){
  for(i in 1:length(pca_grid_clim)){
    for(j in 1:length(pca_grid_clim)){
      if(i<j){
        print(paste(i,"-",j,"/",length(pca_grid_clim)))
        spp1_name = names(pca_grid_clim)[[i]]
        spp1 = pca_grid_clim[[i]]
        spp2_name = names(pca_grid_clim)[[j]]
        spp2 = pca_grid_clim[[j]]
        
        print("Running equivalency test")
        eq.test <- ecospat.niche.equivalency.test_custom(z1=spp1, z2=spp2,
                                                         rep=rep1, alternative = "greater")
        print(paste("Running niche similarity test for",spp1_name,"-",spp2_name))
        sim.test <- ecospat::ecospat.niche.similarity.test(z1=spp1, z2=spp2,
                                                           rep=rep2, alternative = "greater",
                                                           rand.type=2)
        print(paste("Running niche similarity test for",spp2_name,"-",spp1_name))
        sim.test2 <- ecospat::ecospat.niche.similarity.test(z1=spp2, z2=spp1,
                                                            rep=rep2, alternative = "greater",
                                                            rand.type=2)
        
        #pdf(paste("EquivalencyOverlapTests_",species,"_",spp1_name,"_",spp2_name,".pdf",sep=""))
        par(mfrow=c(2,3))
        ecospat::ecospat.plot.overlap.test(eq.test, "D", "Equivalency D")
        ecospat::ecospat.plot.overlap.test(sim.test, "D", paste("Similarity D ",spp1_name,"->",spp2_name,sep=""))
        ecospat::ecospat.plot.overlap.test(sim.test2, "D", paste("Similarity D ",spp2_name,"->",spp1_name,sep=""))
        ecospat::ecospat.plot.overlap.test(eq.test, "I", "Equivalency I")
        ecospat::ecospat.plot.overlap.test(sim.test, "I", paste("Similarity I ",spp1_name,"->",spp2_name,sep=""))
        ecospat::ecospat.plot.overlap.test(sim.test2, "I", paste("Similarity I ",spp2_name,"->",spp1_name,sep=""))
        #dev.off()
        
      }
    }
  }
}


set.seed(100)
pairwiseNicheEquivalence(pca_grid_clim=pca_grid_clim,rep1=100,rep2=1000)


```

With this test, which is again only done on a small subset of the species and a small amount of environmental data, we see that there is support for a higher-than-expected niche equivalency (so the subspecies are more equal in niche than expected by chance) as well as highly similar across species. 


